Static contract checking for Haskell

### 2011-11-08 paper.tex

%% We could use \texttt{true} and \texttt{false} here, to be more
%% consistent with \bad and \unr, but then we should also use
%% \texttt{Nil}, \texttt{Cons}, etc.
\newcommand{\true}{True}
\newcommand{\false}{False}
\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\ensuremath{\mathcal{H}'}}
% Full function application, e.g. $\full{f}(x_1,\ldots,x_n)$. Contrast
% with partial application, e.g. $f~x_1 \ldots x_n$
\newcommand{\full}[1]{\hat{#1}}
\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\renewcommand{\min}[1]{\mbox{min}(#1)}

@@

\begin{figure}
 \begin{center}
  \[  \begin{array}{rcl}
    \dtrans{def} &\to& \phi~(\mbox{formulae}) \\
    \ktrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formulae}) \\
    \ctrans{f \in c} &\to& \phi~(\mbox{formulae}) \\
  \end{array} \]
  \end{center}
  \caption{Translations}
  \label{trans}
\end{figure}

@@

\begin{itemize}
\item $(\phi_\text{projection})$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i~x_1~\dots~x_{a_i}$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\to$: a value $t$ of type $T$\note[nc]{
    ``type'' is used loosely here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

  Note that this is not true for functions: a function is not required
  to use all of its arguments. \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  being crash-free or not doesn't matter.
  \end{itemize}

@@

\begin{figure*}
  $$ \ruleform{\ktrans{data~def} = \Phi} $$
$$\ktrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 \hspace{5.6cm}where
  \begin{center}
    \[  \begin{array}{rcl}
      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{??? $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
      \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = \full{K}_i(\vec{x}) \land a = \full{K}_j(\vec{y})) \\
      \phi_\text{cf} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{a}) \\
      \phi_\text{lazy} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array} \]
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

@@

Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
Figure~\ref{fig:newCTrans}.  To prove that $e \in \{x|p\}$, we get to
assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
we need to prove we're interested in $e$.  The $\min{e}$ constraint
here is not optional, because in addition to concluding 
$e \neq \unr \to p[e/x] \in \{\true,\unr\}$, we also conclude $\min{p[e/x]}$.
The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
$p[e/x] \in \{\true,\unr\}$, and because it can be used to unfold definitions,
we guard it with $\min{e}$.
}
\begin{figure*}
\begin{verbatim}
[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
[[e in {x|p}]]  := e /= unr -> p[e/x] in {true, unr}

@@

\newcommand{\rec}[1]{#1_\text{rec}}
\subsection{Contract checking -- Recursive case}\label{sec:recursiveContracts}
If the function $f=e$ is recursive, then we ask the theorem prover the
following:

>>>>

%% We could use \texttt{true} and \texttt{false} here, to be more
%% consistent with \bad and \unr, but then we should also use
%% \texttt{Nil}, \texttt{Cons}, etc.
\newcommand{\true}{True}
\newcommand{\false}{False}
\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\ensuremath{\mathcal{H}'}}

% Full function application, e.g. $\full{f}(x_1,\ldots,x_n)$. Contrast
% with partial application, e.g. $f~x_1 \ldots x_n$
\newcommand{\full}[1]{\hat{#1}}
% Recursive anonymous versions of functions.
\newcommand{\rec}[1]{#1_\text{rec}}

\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\renewcommand{\min}[1]{\mbox{min}(#1)}

@@

\begin{figure}
 \begin{center}
  \[  \begin{array}{rcl}
    \dtrans{def} &\to& \phi~(\mbox{formula}) \\
    \ktrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formula}) \\
    \ctrans{f \in c} &\to& \phi~(\mbox{formula}) \\
  \end{array} \]
  \end{center}
  \caption{Translations}
  \label{trans}
\end{figure}

@@

\begin{itemize}
\item $(\phi_\text{projection})$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i~x_1~\dots~x_{a_i}$
  on its $k$-th component, $k = 1\text{ to }a_i$.
\item $(\phi_\text{disjoint})$ Distinct constructors have disjoint ranges.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free. \note[nc]{We use the contract
    translation for arrows, and this translation is principled.}
  \item $\to$: a value $t$ of type $T$\note[nc]{
    ``type'' is used loosely here: it means the constructor head is of type $T$.
    } is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the $k$-th argument of $t$ is not crash-free, then the $k$-th
    projection is a crash-free context that throws an expression that
    is not crash-free. \note[nc]{XXX: This translation is ad-hoc.}

  Note that this is not true for functions: a function is not required
  to use all of its arguments. \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  being crash-free or not doesn't matter.

  \note[nc]{Things become more interesting when we add $\min$'s.  For
    the $\gets$ direction, we are asserting that $K_i \in \cfc^{a_i} \to
    \cfc$.  Since we are asserting, it's a minus translation, and so
    the arguments are translated as goals, i.e. we get $\ctrans{x \in
      \cfc}^+$, which is $\min{x} \to \cf{x}$.

  Those $\min$ assumptions on $x$ are very important, as the following
  example illustrates.  Consider proving that addition, defined by
  recursion on the first argument, satisfies $\cfc^2 \to \cfc$. When
  considering $add~(Succ~x)~y$ we need to show
  $\cf{Succ~(\rec{add}~x~y)}$ from $\min{Succ~(\rec{add}~x~y)}$ and
  $\ctrans{\rec{add} \in \cfc^2 \to \cfc}$.  The arrow contract for
  $\rec{add}$ has a $\min{\rec{add}~x~y}$ constraint, but we only know
  $\min$ for the $Succ$ of that.  Hence, the $\min{x}$ assumption,
  from $\ctrans{x \in \cfc}^+$ in $\ctrans{Succ \in \cfc \to
  \cfc}^-$, saves the day.}
  \end{itemize}

@@

\begin{figure*}
  $$ \ruleform{\ktrans{data~def} = \Phi} $$
$$\ktrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{projection} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 \hspace{5.6cm}where
  \begin{center}
    \[  \begin{array}{rcl}
      \phi_\text{projection} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to \bigwedge_{1 \leq j \leq a_i} x_j = sel_{j,K_i}(a) \text{\note[nc]{??? $\min{a}$ or $\min{sel_{j,K_i}{a}}$?}}\\
      \phi_\text{disjoint} &=& \bigwedge_{1 \leq i < j \leq n}~\forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land a = \full{K}_i(\vec{x}) \land a = \full{K}_j(\vec{y})) \\
      \phi_\text{cf} &=& 
        \bigwedge_{1 \leq i \leq n}~\ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
          \land \forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) 
                  \to \cf{a} \to (\bigwedge_{1 \leq j \leq a_i} \cf{x_j}) \\
      \phi_\text{lazy} &=& \bigwedge_{1 \leq i \leq n}~\forall \vec{x},a.~(\min{a} \land a = \full{K}_i(\vec{x})) \to a \neq \unr \land a \neq \bad 
    \end{array} \]
  \end{center}
  \caption{$\ktrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

@@

Finally, consider $\ctrans{e \in \{x|p\}}^+$ and $\ctrans{e \in \{x|p\}}^-$, in
Figure~\ref{fig:newCTrans}.  To prove that $e \in \{x|p\}$, we get to
assume interest in $e$ and $p[e/x]$, whereas, to use $e \in \{x|p\}$,
we need to prove we're interested in $e$.  The $\min{e}$ constraint
here is not optional, because in addition to concluding 
$e \neq \unr \to p[e/x] \in \{\true,\unr\}$, we also conclude $\min{p[e/x]}$.
The $\min{p[e/x]}$ is necessary to unfold definitions in $p[e/x]$ and prove
$p[e/x] \in \{\true,\unr\}$, and because it can be used to unfold definitions,
we guard it with $\min{e}$.
}

\begin{figure*}
\begin{verbatim}
[[e in {x|p}]]+ := min(e) /\ min(p[e/x]) -> [[e in {x|p}]]
[[e in {x|p}]]- := min(e) -> min(p[e/x]) /\ [[e in {x|p}]]
[[e in {x|p}]]  := e /= unr -> p[e/x] in {true, unr}

@@

\subsection{Contract checking -- Recursive case}\label{sec:recursiveContracts}
If the function $f=e$ is recursive, then we ask the theorem prover the
following:
